---
title: "South Bay Waterbird Surveys: Status Update & Evaluation of Monitoring Protocols"
subtitle: "Prepared for California State Coastal Conservancy and Santa Clara Valley Water District"
author: "Max Tarjan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  word_document:
    reference_docx: format.docx
    pandoc_args: ["--smart"]
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, include=FALSE}
#source('Code_waterbird_trend_assessment.R')
figure.num<-0
table.num<-0
```

#BACKGROUND
The South Bay Salt Pond Restoration Project (SBSPRP) is the largest tidal wetland restoration project on the West Coast of the Americas and aims to restore 50-90% of 15,100 acres of historic commercial salt evaporation ponds to tidal marsh in multiple decades. As an adaptive management project, the success of the SBSPRP depends greatly on using the best science to guide current and future management actions such that the resulting total project area supports a diverse species assemblage. In particular, assessments based on extensive monitoring are needed to fulfill SBSPRP Objective 1: “Create, restore, or enhance habitats of sufficient size, function, and appropriate structure to…[m]aintain current migratory bird species that utilize existing salt ponds and associated structures such as levees” (SBSPRP Final Environmental Impact Statement/Report 2007).

##Objectives
SFBBO has been monitoring waterbirds in SBSPRP sites and surrounding Cargill ponds since 2003, and our data are reported annually to the SBSPRP Management Team. Our first task is to create an updated assessment of the status of waterbirds in the context of the SBSPRP’s targets for waterbird populations. 

Our second task is to assess ongoing monitoring efforts to ensure that they are both effective for evaluating SBSPRP targets and efficient to ensure sustainability. Specifically, we aimed to determine an approach that (1) is 50% cheaper than current survey protocols, and (2) can be used to answer the question of whether waterbird populations in the South Bay are above or below targets set by the SBSPRP. We also sought protocols that (1) generate data that are directly comparable to counts from previous years, and (2) allow for simultaneous collection of other data types (e.g. percent of the pond bottom that is vegetated).

#TASK 1 - South Bay Waterbird Population Trends
##Approach
We coordinated with SBSPRP managers through a series of meetings to confirm targets for waterbird guilds and/or species. We then used 15 years of monitoring data and a literature review to assess these targets for migratory waterbirds in the South Bay Salt Pond Restoration Project. We further assessed the targets for nesting waterbirds using SFBBO’s 40-year dataset of colonially nesting waterbirds throughout the Bay Area. All analyses were performed following guidelines of reproducible research so future annual reports will recreate and build on these assessments.

#TASK 2 - Waterbird Survey Protocols
##Approach
We used power analyses and simulations informed by our existing dataset to explore the feasibility of decreasing survey effort using a subsetting protocol. We focused on a subset approach because previous analyses (Tarjan & Heyse 2018) concluded that sampling frequency should remain at current rates (two times in the peak season for a given species) in order to detect ten-year trends in waterbird abundance with sufficient power (beta = 0.80). We 1) researched approaches for decreasing survey effort by subsetting the sites, and 2) evaluated the effect of the subset protocol on our ability to detect trends following methods by Tarjan and Heyse (2018).

###Subsetting Protocols and Trend Detection
The ideal subset of sites would yield counts that are representative of the population. Counts from a subset of sites can become biased if birds shift their locations over time. The potential for this "frame bias" (Bart et al. 2005) leaves researchers unsure of whether a decline occurred due to a decline in the larger population or due to movement away from the surveyed sites (i.e. the frame) over time. A second type of bias called selection bias arises when some sites cannot be surveyed due to access restrictions.

Earlier work by Wood et al. (2010) simulated shorebird counts from sites within the San Francisco Bay Area to evaluate the accuracy and precision of eight subsetting protocols. They tested variations of random sampling, stratification by region such that each region was equally represented in the subsample, and site selection weighted by the abundance of shorebirds in historical data. Based on their assessment of accuracy and precision, they recommended random selection of sites that are weighted using historical data for annual surveys, and stipulated that remaining sites are surveyed every ten years to check for frame bias (i.e. movement of birds between sites). They concluded that with this scheme, survey effort could be reduced by up to 75% and still deliver accurate (+/- 20%) estimates of 20-year population trends at 0.05 and/or 0.15 power.

We applied Wood et al.'s approach with modifications to the geographic scope, species, and detection criteria to identify feasible reductions in survey effort for the SBSPRP.  Previous analyses (Tarjan & Heyse 2018) revealed that survey frequency needs to remain at current levels (twice during the peak season for any given species) to detect Project targets. These results are in agreement with Wood et al.'s findings that decreasing survey effort by 50% (from every one to every two years) dramatically lowers power. An alternative approach to decrease survey effort is to survey fewer locations, which we evaluate here. 

We evaluated our ability to detect trends under sampling schemes where we sample 100%, 50% and 25% of survey sites. Across all subsampling simulations we used a stratified and weighted approach, where the sites (which we define as ponds) were stratified by habitat type and weighted by historical bird abudance. The birds within the SBSPRP represent diverse guilds (e.g. shorebirds, dabbling ducks, diving ducks) so their habitat preferences differ and high density areas are not necessarily representative of the preferred area for all species. To incorporate representative sites for all guilds, we randomly selected sites weighted by density for each guild. We then randomized the order of guilds and selected the top sites in sequential order for each guild until the target number of sites was reached. 

...habitat type = managed pond or restored tidal marsh...weighting takes into account multiple species...adapts based on which species are already covered?? somewhat stochastic weights??

###Survey Cost
```{r, echo=FALSE}
cost.goal<-50000
new.staff.hrs<-round((cost.goal - (91144.12-(2377.25+394)*0.7*mean(c(17.77*1.284,24.62*1.284))))/mean(c(17.77*1.284,24.62*1.284)),0)
new.staff.hrs.cost<-round(cost.goal - (91144.12-(2377.25+394)*0.7*mean(c(17.77*1.284,24.62*1.284))),2)
```

The three-year average of annual spending for 2015-2017 is \$84371.46. We sought a survey protocol that could be implemented on an annual budget of \$`r cost.goal`, which represents a `r round((84371.46-cost.goal)/84371.46*100,0)`% decrease in survey costs. See Appendix I for a previous project budget. Reductions in budget items will depend on how the protocol changes are implemented. We expect that the following costs would remain the same with a new survey protocol, assuming the protocol was implemented with the same frequency: data analysis, data management, meetings, project management, and report writing, as well as supplies, mileage and tolls, and conferences. Reduced costs would be reflected in fewer staff hours spent on data entry and field work. Past data entry required 394 hours, while field work required 2377.25 hours each year. Roughly 30% of these hours can be fulfilled by interns. The rate for interns will remain at a minimum of $500 per month. Achieving the goal for cost decreases requires that staff hours on field work and data entry are decreased to `r new.staff.hrs` total hours, which would cost \$`r new.staff.hrs.cost` with overhead. Maintaining the ratio of field work hours to data entry hours indicates that staff would need to spend `r new.staff.hrs*0.858` hours on fieldwork and `r new.staff.hrs*0.142` hours on data entry, which represents a `r round(((2377.25+394)-new.staff.hrs)/(2377.25+394)*100,0)`% reduction in field work hours.

#WORKS CITED
Tarjan, L.M. & V. Heyse. 2018. Evaluation of Waterbird Monitoring Protocols for the South Bay Salt Pond Restoration Project. Report prepared for the South Bay Salt Pond Restoration Project Management Team.

Wood, J., Page, G., Reiter, M., Liu, L., & Robinson-nilsen, C. (2010). Abundance and Distribution of Wintering Shorebirds in San Francisco Bay , 1990-2008 : Population Change and Informing Future Monitoring. Petaluma, CA.
